United Monoids

### 2022-01-12 paper/main.tex
Another proofreading pass

\emph{Context:} Graphs and various graph-like combinatorial structures, such as
preorders and hypergraphs, are ubiquitous in programming. This paper focuses on
representing graphs in a purely functional programming language like Haskell.
There are several existing approaches; one of the most recently developed ones
is the ``algebraic graphs'' approach (2017). It uses an algebraic data type to
represent graphs and has attracted users, including from industry, due to its
emphasis on equational reasoning and making a common class of bugs impossible
by eliminating internal invariants.

\emph{Inquiry:} The previous formulation of algebraic graphs did not support
edge labels, which was a serious practical limitation. This paper redesigns the
main algebraic data type and removes this limitation.

\emph{Approach:} We follow a fairly standard approach of parameterising a data
structure with a semiring of edge labels. The newly developed data structure is
both more general and simpler: the two ways of composing graphs used in the
original formulation can now be obtained from a single operation by fixing the
semiring parameter to zero and one, respectively.

\emph{Knowledge:} By analysing the resulting data structure and formulating laws
for interpreting expressions of the corresponding type, we discover an unusual
algebraic structure, which we call ``united monoids'', that is, a pair of
monoids whose unit elements coincide. We believe that it is worth studying
united monoids in their full generality, going beyond the graphs which prompted
their discovery. To that end, we characterise united monoids with a minimal set
of axioms, prove a few basic theorems, and discuss several notable examples.

\emph{Grounding}: We validate the presented approach by implementing it in the
open-source \textsf{algebraic-graphs} library. Our theoretical contributions
are supported by proofs that are included in the paper itself and are also
available in a machine-checked artifact.

\emph{Importance:} By extending algebraic graphs with support for edge labels,
we make them suitable for a much larger class of possible applications. By
studying united monoids, we provide a theoretical foundation for further
research in this area.
\end{abstract}

@@

\noindent
For example, to compute the \emph{size} of a tree, i.e. the number of leaves in
it, we can fold the tree into an \code{Int} as follows.

@@

\vspace{-1mm}
\begin{itemize}
  \item The lack of connectivity is denoted by~\zero. For instance, zero network
        bandwidth.
  \item Addition~$\add$ is a commutative\footnote{
            An operation $\bullet$ is \emph{commutative} if
            $a \bullet b = b \bullet a$.
        } operation for combining connections in \emph{parallel}; e.g.,
        selecting the \emph{maximum} bandwidth link between two locations.
  \item Multiplication~$\mul$ is for combining connections in \emph{sequence}.
        Continuing the example, $\mul$ corresponds to taking the \emph{minimum}
        bandwidth of network links on route between two locations.
        Multiplication distributes\footnote{
            \emph{Distributivity} of $\otimes$ over $\oplus$ means
            $a \otimes (b \oplus c) = (a \otimes b) \oplus (a \otimes c)$ and
            $(a \oplus b) \otimes c = (a \otimes c) \oplus (b \otimes c)$.
        } over addition and has \zero as a zero\footnote{
            An element $z$ is a \emph{zero} (or an \emph{annihilating element})
            of an operation $\bullet$ if $z \bullet a = a \bullet z = z$.
        }.
  \item Finally, \one denotes either a unit of connectivity or infinite
        connectivity, depending on a specific semiring. In the bandwidth
        semiring, we have $\text{\one} = \infty$.
\end{itemize}

@@

\begin{itemize}
    \item $\text{\hcode{node zero a b}} = \text{\hcode{node zero b a}}$, i.e.,
          commutativity of \zero-labelled nodes;
    \item $\text{\hcode{node zero a a}} = \text{\hcode{a}}$, i.e.,
    idempotence\footnote{
         An operation $\bullet$ is \emph{idempotent} if $a \bullet a = a$.
    } of \zero-labelled nodes.
\end{itemize}

@@

When working with simplicial sets, we will extend the set of requirements on
interpreting trees by adding the \emph{containment} (see~\S\ref{sec-united}) and
distributivity requirements:

@@

          \vspace{0.5mm}\hspace{10mm}$a \arr (b \dia c) = (a \arr b) \dia (a \arr c)$~~and~~$(a \dia b) \arr c = (a \arr c) \dia (b \arr c)$.
\end{itemize}

@@

\noindent
Now that we know the rules of interpreting trees over the Boolean semiring, we
need to choose a target representation for simplicial sets. Alas, there is no
standard \code{Data.SimplicialS}\code{et}, so we need to come up with our own.
Here is a simple candidate:

@@

\begin{lstlisting}
vertex :: a -> SimplicialSet a
vertex = Set.singleton . NonEmpty.singleton
!\vspace{-1mm}!
overlay :: Ord a => SimplicialSet a -> SimplicialSet a -> SimplicialSet a
overlay = Set.union
!\vspace{-1mm}!
connect :: Ord a => SimplicialSet a -> SimplicialSet a -> SimplicialSet a
connect x y = Set.unions [x, y, Set.map (uncurry (!\hdia!)) (Set.cartesianProduct x y)]
!\vspace{-1mm}!
toSimplicialSet :: Ord a => Tree Bool a -> SimplicialSet a
toSimplicialSet = fold vertex (\s -> if s then connect else overlay)
\end{lstlisting}

@@

\begin{itemize}
    \item Decomposition:
          $a \arr b \arr c = (a \arr b) \dia (a \arr c) \dia (b \arr c)$.
\end{itemize}

@@

\begin{lstlisting}
connect :: Ord a => Graph a -> Graph a -> Graph a
connect (v1,e1) (v2,e2) = (Set.union v1 v2, Set.unions [e1, e2, Set.cartesianProduct v1 v2])
\end{lstlisting}

\noindent
It is not obvious that this interpretation respects the decomposition
requirement, so let us check it. For the vertex set, the requirement holds
because $\cup$ is idempotent:

@@

\begin{itemize}
    \item For \emph{undirected graphs}, $\arr$ must commute for leaves:
    \hcode{leaf}$~a \arr $\hcode{leaf}~$b =~$\hcode{leaf}$~b \arr $\hcode{leaf}~$a$.
    Taken with other requirements, this leads to general commutativity
    $a \arr b = b \arr a$.
    \item For \emph{reflexive graphs}, $\arr$ must have no effect on leaves:
    \hcode{leaf}$~a =~$\hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$a$.
    \item For \emph{bipartite graphs}, $\arr$ must have no effect on leaves
    corresponding to vertices from the same part. That is,
    $\forall (a,b) \in (P \times P) \cup (Q \times Q),\ $\hcode{leaf}$~a\,\dia\,$\hcode{leaf}~$b =~$\hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$b$,
    where $P$~and~$Q$ are the two parts of the bipartite graph.
    \item For \emph{transitively-closed graphs}, see \S\ref{sec-preorder}.
\end{itemize}

@@

% In this section we will see how to interpret trees as preorders. We will first
% build on the infrastructure developed for graphs in~\S\ref{sec-graph}, and then
% discuss a generalisation.
\noindent
As mentioned earlier in~\S\ref{sec-other-graphs}, reflexivity can be expressed
as a new tree folding requirement, which states that connecting a leaf to itself
must be redundant:

@@

\noindent
To ensure that preorder expressions are interpreted
``modulo transitive connections'', we also add the following transitivity
requirement. It states that adding/removing a transitive connection $a \arr c$
to/from the tree $a \arr b \dia b \arr c$ must be redundant:

@@

% or $a \arr b\ \dia\ b \arr c = a \arr b \arr c$
% Note that the alternative (second) definition makes use of the decomposition
% law.
\noindent
Reading the reflexivity and transitivity requirements from left to right, one
can think of them as introducing the preorder terms $a \preceq a$ and
$a \preceq c$ under suitable assumptions.
% Conversely, reading the requirements from right to left corresponds to
% eliminating the terms $a \preceq a$ and $a \preceq c$, or simplifying them away.

@@

\noindent
Then generalised reflexivity and transitivity requirements can be expressed as follows.

@@

The \hcode{Tree} data type cannot rule out cyclic terms like
$a \arr b \dia b \arr a$ at the type level. If these terms are undesirable, we
can turn them into \emph{errors} by adding two requirements:

\begin{itemize}
    \item \emph{Cycle}:
    \hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$a = \bot$, where $\bot$ is a
    ``cycle error'' value in the target type.
    \item \emph{Cycle error is a zero}: $a \dia \bot = a \arr \bot = \bot \arr a = \bot$.
\end{itemize}

@@

\noindent
To give meaning to parallel edges, we introduce the \emph{parallel composition}
requirement, which is similar to the sequential composition
from~\S\ref{sec-preorder} but concerns \zero and $\add$~instead:

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a \arr b\ \dia\ a\ \dia\ b & = & a \xrightarrow{\one} b\ \dia\ a \xrightarrow{\zero} b & \text{(definitions of $\arr$, $\dia$ and $a \xrightarrow{x} b$)}\\
 & = & a \xrightarrow{\one\,\raisebox{0.15mm}{\add}\,\zero} b & \text{(parallel composition)}\\
 & = & a \xrightarrow{\one} b & \text{(unit of $\add$)}\\
 & = & a \arr b & \text{(definitions of $\arr$ and $a \xrightarrow{x} b$)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
Another tool that we need for interpreting edge-labelled graphs is the
generalised decomposition requirement, which gives meaning to nested nodes.

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
(a \dia b) \xrightarrow{x} c &\!\!=\!\!& (a \xrightarrow{\zero} b) \xrightarrow{x} c &\!\!\text{(definitions of $\dia$ and $a \xrightarrow{x} b$)}\\
 &\!\!=\!\!& a \xrightarrow{\zero} b\ \dia\ a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(left decomposition)}\\
 &\!\!=\!\!& a\ \dia\ b\ \dia\ a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(definitions of $\dia$ and $a \xrightarrow{x} b$)}\\
 &\!\!=\!\!& (a\,\dia\,a\,\xrightarrow{x}\,c)\ \dia\ (b\,\dia\,b\,\xrightarrow{x}\,c) &\!\!\text{(associativity and commutativity of \dia)}\\
 &\!\!=\!\!& a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(containment)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
To conclude, the set of requirements for interpreting edge-labelled graphs are:
parallel composition, decomposition, and commutativity of $\dia$ inherited from
sets~(\S\ref{sec-set}).

@@

overlay :: (Ord a, Semiring s) => LGraph s a -> LGraph s a -> LGraph s a
overlay = Map.unionWith (Map.unionWith (!\h{\hadd}!)) -- Parallel composition of edge labels
\end{lstlisting}

@@

\begin{lstlisting}
connect :: (Ord a, Semiring s, Eq s) => s -> LGraph s a -> LGraph s a -> LGraph s a
connect s x y
    | s == zero = overlay x y -- Filter out redundant !\h{zero}!-labelled edges
    | otherwise = Map.unionsWith (Map.unionWith (!\h{\hadd}!)) [ x, y, newEdges ]
  where
    outgoing = Map.fromSet (const s) (Map.keysSet y)
    newEdges = Map.fromSet (const outgoing) (Map.keysSet x)
\end{lstlisting}

@@

\noindent
It is useful to combine edge-labelled graphs with the sequential composition
requirement described in the previous section~(\S\ref{sec-preorder}). By
computing the reflexive and transitive closure of an edge-labelled graph, it is
possible to solve a variety of semiring optimisation problems, e.g.,
see~\cite{2013_semirings_dolan}\cite{mohri2002semiring}.

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
z^{+} & = & z^{\cdot}\ +\ z^{+} & \text{($z^{+}$ is a zero of $+$)}\\
 & = & z^{\cdot} \cdot z^{+}\ +\ z^{+} & \text{($z^{\cdot}$ is a zero of $\cdot$)}\\
 & = & z^{\cdot} \cdot z^{+} & \text{(containment)}\\
 & = & z^{\cdot} & \text{($z^{\cdot}$ is a zero of $\cdot$)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
Back in~\ref{sec-acyclic-graphs}, we used zero $\bot$ to model ``cycle errors''.
As we now see, requiring both $a \dia \bot = a$ and
$a \arr \bot = \bot \arr a = \bot$ was unnecessary: we could have picked just
one of these requirements and the other one would follow.

@@

Various flavours of parallel and sequential composition often form united
monoids. In the paper on Concurrent Kleene Algebra~\cite{hoare2011concurrent},
Hoare et al. use the term ``bimonoid'' to refer to such structures (without
investigating them in detail). In category theory, ``bimonoid'' is used for an
unrelated concept (a structure with a monoid and a
comonoid~\cite{porst_bimonoids}). The author of this paper chose to use the term
``united monoids'', which appears to be unused and is also more specific
compared to ``bimonoid''.

@@

\acks
I would like to thank everyone who contributed to this research by giving
feedback on earlier versions of this work, participating in numerous online and
in-person discussions, and last but not least, expressing genuine interest and
encouragement.
Here is a most likely incomplete list (I apologise for any omissions):
Arseniy Alekseyev, Roland Backhouse, Gershom Bazerman,
Dave Clarke, Ulan Degenbaev, Chris Doran, Victor Khomenko, Adithya Kumar,
Dave Long, Anton Lorenzen, Georgy Lukyanov, Alp Mestanogullari, Alexandre Moine,
Andreas Nuyts, Greg Pfeil, Stefan Plantikow, Armando Santos, Danil Sokolov,
Sjoerd Visscher, Leo White and Alex Yakovlev. I am also grateful to the
Programming Journal's Programme Committee for the thorough reviews that helped
me improve the final version of the paper.

>>>>

Graphs and various graph-like combinatorial structures, such as
preorders and hypergraphs, are ubiquitous in programming. This paper focuses on
representing graphs in a purely functional programming language like Haskell.
There are several existing approaches; one of the most recently developed ones
is the ``algebraic graphs'' approach (2017). It uses an algebraic data type to
represent graphs and has attracted users, including from industry, due to its
emphasis on equational reasoning and making a common class of bugs impossible
by eliminating internal invariants.

The previous formulation of algebraic graphs did not support edge labels, which
was a serious practical limitation. In this paper, we redesign the main
algebraic data type and remove this limitation. We follow a fairly standard
approach of parameterising a data structure with a semiring of edge labels. The
new formulation is both more general and simpler: the two operations for
composing graphs used in the previous work can now be obtained from a single
operation by fixing the semiring parameter to zero and one, respectively.

By instantiating the new data type with different semirings, and working out
laws for interpreting the resulting expression trees, we discover an unusual
algebraic structure, which we call ``united monoids'', that
is, a pair of monoids whose unit elements coincide. We believe that it is worth
studying united monoids in their full generality, going beyond the graphs which
prompted their discovery. To that end, we characterise united monoids with a
minimal set of axioms, prove a few basic theorems, and discuss several notable
examples.

We validate the presented approach by implementing it in the
open-source \textsf{algebraic-graphs} library. Our theoretical contributions
are supported by proofs that are included in the paper and have also
been machine-checked in Agda. By extending algebraic graphs with support for
edge labels, we make them suitable for a much larger class of possible
applications. By studying united monoids, we provide a theoretical foundation
for further research in this area.
\end{abstract}

@@

\noindent
For example, to compute the \emph{size} of a tree, i.e., the number of leaves in
it, we can fold the tree into an \code{Int} as follows.

@@

\vspace{-1mm}
\begin{itemize}
  \item The lack of connectivity is denoted by~\zero. For instance, zero network
        bandwidth.
  \item Addition~$\add$ is a commutative\footnote{
            An operation $\bullet$ is \emph{commutative} if
            $a \bullet b = b \bullet a$.
        } operation for combining connections in \emph{parallel}; e.g.,
        selecting a link with the \emph{maximum} bandwidth from available
        alternatives.
  \item Multiplication~$\mul$ combines connections in \emph{sequence}.
        Continuing the example, to compute the bandwidth of a route, we take the
        \emph{minimum} over all links that need to be traversed. Multiplication
        distributes\footnote{
            \emph{Distributivity} of $\otimes$ over $\oplus$ means
            $a \otimes (b \oplus c) = (a \otimes b) \oplus (a \otimes c)$ and
            $(a \oplus b) \otimes c = (a \otimes c) \oplus (b \otimes c)$.
        } over addition and has \zero as its zero\footnote{
            An element $z$ is the \emph{zero} (or the
            \emph{annihilating element}) of an operation $\bullet$ if
            $z \bullet a = a \bullet z = z$.
        }.
  \item Finally, \one denotes either a unit of connectivity or infinite
        connectivity, depending on a specific semiring. In the bandwidth
        semiring, we have $\text{\one} = \infty$.
\end{itemize}

@@

\begin{itemize}
    \item $\text{\hcode{node zero}}~a~b = \text{\hcode{node zero}}~b~a$, i.e.,
          commutativity of \zero-labelled nodes;
    \item $\text{\hcode{node zero}}~a~a = a$, i.e.,
    idempotence\footnote{
         An operation $\bullet$ is \emph{idempotent} if $a \bullet a = a$.
    } of \zero-labelled nodes.
\end{itemize}

@@

When working with simplicial sets, we will extend the set of requirements on
interpreting trees by adding the \emph{containment} (see~\S\ref{sec-united}) and
distributivity laws:

@@

          \vspace{0.5mm}\hspace{10mm}$a \arr (b \dia c) = (a \arr b)\,\dia\,(a \arr c)$~~and~~$(a \dia b) \arr c = (a \arr c)\,\dia\,(b \arr c)$.
\end{itemize}

@@

\noindent
Now that we know the laws for interpreting trees over the Boolean semiring, we
need to choose a target representation for simplicial sets. Alas, there is no
standard \code{Data.SimplicialS}\code{et}, so we need to come up with our own.
Here is a simple candidate:

@@

\begin{lstlisting}
vertex :: a -> SimplicialSet a
vertex = Set.singleton . NonEmpty.singleton
!\vspace{-1mm}!
overlay :: Ord a => SimplicialSet a -> SimplicialSet a -> SimplicialSet a
overlay = Set.union
!\vspace{-1mm}!
connect :: Ord a => SimplicialSet a -> SimplicialSet a -> SimplicialSet a
connect x y =
    Set.unions [x, y, Set.map (uncurry NonEmpty.append) (Set.cartesianProduct x y)]
!\vspace{-1mm}!
toSimplicialSet :: Ord a => Tree Bool a -> SimplicialSet a
toSimplicialSet = fold vertex (\s -> if s then connect else overlay)
\end{lstlisting}

@@

\begin{itemize}
    \item Decomposition:
          $a \arr b \arr c = (a \arr b)\,\dia\,(a \arr c)\,\dia\,(b \arr c)$.
\end{itemize}

@@

\begin{lstlisting}
connect :: Ord a => Graph a -> Graph a -> Graph a
connect (v1,e1) (v2,e2) = (Set.union!\,!v1!\,!v2, Set.unions!\,![e1,!\,!e2,!\,!Set.cartesianProduct!\,!v1!\,!v2])
\end{lstlisting}

\noindent
It is not obvious that this interpretation respects the decomposition law, so
let us check it. For the vertex set, the law holds because $\cup$ is idempotent:

@@

\begin{itemize}
    \item For \emph{undirected graphs}, $\arr$ must commute for leaves:
    \hcode{leaf}$~a \arr $\hcode{leaf}~$b =~$\hcode{leaf}$~b \arr $\hcode{leaf}~$a$.
    Taken with other requirements, this leads to general commutativity
    $a \arr b = b \arr a$.
    \item For \emph{reflexive graphs}, $\arr$ must have no effect on leaves:
    \hcode{leaf}$~a =~$\hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$a$.
    \item For \emph{bipartite graphs}, $\arr$ must have no effect on leaves
    corresponding to vertices from the same part. That is,
    $\forall (a,b) \in (P \times P) \cup (Q \times Q),\ $\hcode{leaf}$~a\,\dia\,$\hcode{leaf}~$b =~$\hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$b$,
    where $P$~and~$Q$ are the two parts of the bipartite graph.
    \item For \emph{transitively-closed} and \emph{acyclic graphs},
    see~\S\ref{sec-preorder}.
\end{itemize}

@@

% In this section we will see how to interpret trees as preorders. We will first
% build on the infrastructure developed for graphs in~\S\ref{sec-graph}, and then
% discuss a generalisation.
\noindent
As mentioned earlier in~\S\ref{sec-other-graphs}, reflexivity can be expressed
as a new law, which requires that connecting a leaf to itself must be redundant:

@@

\noindent
To ensure that preorder expressions are interpreted
``modulo transitive connections'', we also add the following transitivity law.
It states that adding/removing a transitive connection $a \arr c$ to/from the
tree $a \arr b \dia b \arr c$ must be redundant:

@@

% or $a \arr b\ \dia\ b \arr c = a \arr b \arr c$
% Note that the alternative (second) definition makes use of the decomposition
% law.
\noindent
Reading the reflexivity and transitivity laws from left to right, one can think
of them as introducing the preorder terms $a \preceq a$ and $a \preceq c$ under
suitable assumptions.
% Conversely, reading the requirements from right to left corresponds to
% eliminating the terms $a \preceq a$ and $a \preceq c$, or simplifying them away.

@@

\noindent
Then generalised reflexivity and transitivity laws can be expressed as follows.

@@

The \hcode{Tree} data type cannot rule out cyclic terms like
$a \arr b \dia b \arr a$ at the type level. If these terms are undesirable, we
can turn them all to \emph{errors} by adding two new laws:

\begin{itemize}
    \item \emph{Cycle}:
    \hcode{leaf}$~a\,\arr\,$\hcode{leaf}~$a = \bot$, where $\bot$ is a
    ``cycle error'' value in the target type.
    \item \emph{Zero}: $a \dia \bot = a \arr \bot = \bot \arr a = \bot$.
\end{itemize}

@@

\noindent
To give meaning to parallel edges, we introduce the \emph{parallel composition}
law. It is similar to the sequential composition from~\S\ref{sec-preorder}
but concerns \zero and $\add$~instead of~\one~and~$\mul$:

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a \arr b\ \dia\ a\ \dia\ b & = & a \xrightarrow{\one} b\ \dia\ a \xrightarrow{\zero} b & \text{(definitions of $\arr$, $\dia$ and $\xrightarrow{x}$)}\\
 & = & a \xrightarrow{\one\,\raisebox{0.15mm}{\add}\,\zero} b & \text{(parallel composition)}\\
 & = & a \xrightarrow{\one} b & \text{(unit of $\add$)}\\
 & = & a \arr b & \text{(definitions of $\arr$ and $\xrightarrow{x}$)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
Another tool that we need for interpreting edge-labelled graphs is the
generalised decomposition law, which gives meaning to nested nodes.

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
(a \dia b) \xrightarrow{x} c &\!\!=\!\!& (a \xrightarrow{\zero} b) \xrightarrow{x} c &\!\!\text{(definitions of $\dia$ and $\xrightarrow{x}$)}\\
 &\!\!=\!\!& a \xrightarrow{\zero} b\ \dia\ a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(left decomposition)}\\
 &\!\!=\!\!& a\ \dia\ b\ \dia\ a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(definitions of $\dia$ and $\xrightarrow{x}$)}\\
 &\!\!=\!\!& (a\,\dia\,a\,\xrightarrow{x}\,c)\ \dia\ (b\,\dia\,b\,\xrightarrow{x}\,c) &\!\!\text{(associativity and commutativity of \dia)}\\
 &\!\!=\!\!& a \xrightarrow{x} c\ \dia\ b \xrightarrow{x} c &\!\!\text{(containment)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
To conclude, the complete set of requirements for interpreting edge-labelled
graphs is: commutativity of $\dia$ inherited from sets~(\S\ref{sec-set}),
parallel composition and decomposition.

@@

overlay :: (Ord a, Semiring s) => LGraph s a -> LGraph s a -> LGraph s a
overlay = Map.unionWith (Map.unionWith (!\h{\hadd}!)) -- Parallel composition
\end{lstlisting}

@@

\begin{lstlisting}
connect :: (Ord a, Semiring s, Eq s) => s -> LGraph s a -> LGraph s a -> LGraph s a
connect s x y
    | s == zero = overlay x y -- Filter out redundant !\h{zero}!-labelled edges
    | otherwise = Map.unionsWith (Map.unionWith (!\h{\hadd}!)) [ x, y, newEdges ]
  where
    outgoing = Map.fromSet (const s) (Map.keysSet y)
    newEdges = Map.fromSet (const outgoing) (Map.keysSet x) -- Decomposition
\end{lstlisting}

@@

\noindent
It is useful to combine edge-labelled graphs with the sequential composition
law described in the previous section~(\S\ref{sec-preorder}). By computing the
reflexive and transitive closure of an edge-labelled graph, it is possible to
solve a variety of semiring optimisation problems, e.g.,
see~\cite{2013_semirings_dolan}\cite{mohri2002semiring}.

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
z^{+} & = & z^{\cdot}\ +\ z^{+} & \text{($z^{+}$ is the zero of $+$)}\\
 & = & z^{\cdot} \cdot z^{+}\ +\ z^{+} & \text{($z^{\cdot}$ is the zero of $\cdot$)}\\
 & = & z^{\cdot} \cdot z^{+} & \text{(containment)}\\
 & = & z^{\cdot} & \text{($z^{\cdot}$ is the zero of $\cdot$)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
Back in~\ref{sec-acyclic-graphs}, we used zero $\bot$ to model ``cycle errors''.
As we now see, requiring both $a \dia \bot = a$ and
$a \arr \bot = \bot \arr a = \bot$ was unnecessary: we could have picked just
one of these laws and the other one would follow.
For general (i.e., not acyclic) graphs, the fully-connected graph acts as the
zero.

@@

Various flavours of parallel and sequential composition often form united
monoids. In the paper on Concurrent Kleene Algebra~\cite{hoare2011concurrent},
Hoare et al. use the term ``bimonoid'' to refer to such structures (without
investigating them in detail). In category theory, ``bimonoid'' is used for an
unrelated concept (a structure with a monoid and a
comonoid~\cite{porst_bimonoids}). The author of this paper chose to use the term
``united monoids'', because it appears to be unused and is also more specific
compared to ``bimonoid''.

@@

\acks
I would like to thank everyone who contributed to this research by giving
feedback on earlier versions of this work (particularly,~\cite{mokhov_alga}
and~\cite{united_monoids_blog}), participating in numerous online and
in-person discussions, and last but not least, expressing genuine interest and
encouragement.
Here is a most likely incomplete list (I apologise for any omissions):
Arseniy Alekseyev, Roland Backhouse, Gershom Bazerman, Dave Clarke,
Ulan Degenbaev, Dominique Devriese, Chris Doran, Victor Khomenko, Adithya Kumar,
Dave Long, Anton Lorenzen, Georgy Lukyanov, Alp Mestanogullari, Alexandre Moine,
Andreas Nuyts, Greg Pfeil, Stefan Plantikow, Armando Santos, Danil Sokolov,
Sjoerd Visscher, Leo White and Alex Yakovlev. I am also grateful to the
Programming Journal's Programme Committee for the thorough reviews that helped
me improve the final version of the paper.
